A game-theory modeling approach to utility and strength of interactions dynamics in biomedical research social networks

### 2016-11-19 manuscript_3.tex
interaction strength needs less interpretation
attachment and trust are terms for different phenomena

  What would happen if researchers in a community were given the same
  amount of resources and then set free to interact among each other?
  How would these resources be distributed in the population after
  many interactions? The final result may be determined by the
  researchers' social and political power, an by their
  prestige. Nevertheless, all these attributes depend, to a certain
  degree, on the research collaboration social network in which
  scientists are embedded. Here we develop a model based on the
  Prisoner's Dilemma implemented on a collaboration network. With this
  model we explore the distribution of resources and the dynamics of
  the strengthening or weakening of collaboration connections among
  partners, as scientists cooperate or defect in order to gain access
  to others' resources. The network is based on data about projects
  that were awarded with a grant. In this sense, we assume that
  resources are funds, equipment and time. We define Fitness as how
  much access a researcher has to the resources of his collegues.  We
  tested our simulation on the real biomedical research network and
  compared the results with an Erd\"{o}s-Reny\'i, a Watts-Strogatz
  small-world and Barab\'asi-Albert topologies. Different topologies
  display different fitness and connections strength
  distributions. Moreover, the distribution of fitness and connections
  strength in the researchers network is similar to that of
  Barab\'asi-Albert and Watts-Strogatz topologies, respectively. We
  believe that fitness distribution in the researchers network
  suggests that there are socio-cultural mechanisms governing the
  network that produce an asymmetric distribution of resources. The
  high distribution of strong connections might reflect some sort of
  subordination among researchers by which they are morally obliged to
  cooperate by the same socio-cultural mechanisms. The range around
  the threshold that regulates the decision to cooperate or defect
  according to the agent's historical balance between utility and
  strength of collaborative relationships and carrying capacity of the
  system is small, suggesting that there is a region in which a phase
  transition takes place from a population of cooperators to a
  population of defectors. Simulations like this may help to develop
  science policies to promote fair distribution of resources.

@@

\textbf{Notes:}

\begin{itemize}

%% \item In this paper there are two games played simultaneously: One for
%% maximazing individual utility; the second for maximazing the connection
%% strength. We are interested in how they affect eachother in the context of a
%% network of collaboration.

\item We change utility for fitness. Utility seems a better concept in order to
  reflect players' preferences.

%% \item In the context of our paper, utility represents
%%   the preference for having access to the resources of others. Maximazing
%%   utility means that the player is increasing its access to the resources of
%%   other players.  Utility function is just the sum of the payoffs of playing
%%   with its neighbors. As for the payoff matrix, the best strategy for a player
%%   is to defect. Any other option will not be an optimal payoff. The opossing
%%   force is in the other game. The game for maximazing the strength of the
%%   connection among players. In this case defecting or cooperating? is the best
%%   strategy as it pays the most when coordinating regardeless they both cooperate
%%   or defect, this is because when both cooperate the interaction gets a positive
%%   payoff, when both defect, the interaction doesn't get affected; but if they
%%   anti-coordinate, then the interaction looses. In this sense, this might be a
%%   game were there are multiple (i.e., 2 points or strategies) equilibria. It
%%   seems that there is no single dominant strategy.}\\

\item Choosing a strategy depends on the balance $\eta'$ equation --look at page
  9. This
  equation states that the probability of an agent to cooperate increases as: a)
  its utility increases --as it moves away from 0, or; b) The strength of the
  connection with its peers increases --as it moves away from 0, or; c) both
  increases.

\item{En la pagina 9 decimos que ``In order for the agent to choose to cooperate
  or defect, a degree of willingness to cooperate is assigned to each agent.'' No estoy
  seguro que ``degree of willingness to cooperate'' sea la mejor manera de
  decirlo, mi duda está en la palabra ``degree'', por lo que cambi\'e la
  oraci\'on a: The probability for an agent to cooperate or defect depends on a
number that referes to a historical balance between average utility and the average strength of
the connections with its neigbors. }

\end{itemize}

@@

 {\color{red}In this article we explore the network effect on the distribution
   of players having access to certain amount of resources from other players in the
   network and the distribution of the strength of connections among
   them. Particularly, we implemented two games played simultaneously: 
   one for maximazing individual utility based on the iterated Prisoner's
   Dilemma; the other, a coordination game for maximazing the connection strength between
   players. We are interested in how they affect each other in 
   the context of a network of scientific collaboration under the idea that while
   researchers are interested in maximazing their utilities, they also know that
   it is important to invest in building collaborative relationships. These two
   behaviors are explored in a biomedical research community of M\'exico.}\\

{\color{red}In the context of our paper, utility represents
  the preference for having access to the resources of others. Maximazing
  utility means that the player is increasing its access to the resources of
  other players. The value of the Utility function for a player is the sum of the payoffs of playing
  with his neighbors. The opossing force comes from the other concurrent game: players
  trying to maximize the strength of their connections to other players. In the coordinating game
  the best strategy is to adopt the same strategy as the other player, as it pays the most
  regardless of cooperation or defection in the utility game. When
  both cooperate the interaction gets a positive payoff, when both defect, the
  interaction doesn't get affected; but if they anti-coordinate, then the
  interaction looses. Finally, cooperation is a central feature of scientific work.
  For our biomedical network, cooperation can be thought of
  as sharing resources such as time, students, equipment, even money. Examples of defection
  to a cooperator are ghost authorship or
  prestige authorship.

  % este ejemplo de defection no es nada claro:
  %
  % Another example may be a working strategy in which some
  % members of a team do the work at certain moment, and then another part 
  % does its part in some other moment in the future --some sort of
  % reciprocity.

@@

The manuscript is structured in five sections. First we describe
\textit{FOSISS}, the main program for grants destined to biomedical applied
research in M\'exico. This is the source of the
database from which we created the researchers collaboration network. Next we
describe our model
and the different network topologies on which we explored it.
We then present our results and discuss them briefly. In the last section
we draw some final remarks and conclusions.

@@

\emph{FOSISS} or Sectoral Fund for Health and Social Security Research
(\emph{Fondo Sectorial en Investigaci\'on en Salud y Seguridad Social}) is one of
such funds. FOSISS is constituted by CONACyT, SSA, IMSS and ISSSTE,\footnote{SSA
  is the acronym for Secretariat of Health \emph{Secretar\'ia de Salud}; IMSS is
  the acronym for Social Security Mexican Institute (\emph{Instituto Mexicano
    del Seguro Social}); ISSSTE stands for Institute for Social Security and
  Services for State Workers (\emph{Instituto de Seguridad y Servicios Sociales
    de los Trabajadores del Estado})} being all of them the major public health
providers and research institutions in the country. Every year CONACyT opens a
call for funds limited to a set of health research areas previously defined by a
group of experts. Such areas range from public health issues to chronic and
degenerative diseases.\\

Eligibility is open to public and private health research sectors,
however most applicants are public universities and research
institutions. From 2002 through 2013, there were 91 institutions funded that
comprised 4988 researchers.\\

From these data some important considerations should be made clear. The
population represented in the data include principal investigators (PIs),
associate researchers, postdoctoral associates, postgraduate and undergraduate
students.
% wut? which information is not specified? which is pi and which is undergrad is not specified?
Unfortunately, this information is not specified in the
database. Still, this is something we acknowledge and it's important because
researchers in our network are under different circumstances and we know
that this diversity has a real impact on the structure and eventually on the
dynamics of the network, as well as on the results of our model.\\

Our database includes the
name of the project, the year it was approved for funding and the research area
to which it was assigned. It specifies the names of PIs or the people 
responsible for the project and the names of collaborators.
Researchers can be PIs in one project and collaborators on a different project. The institutional
affiliation of all participants is included. Through this affiliation
we know the principal institution
behind every project.\\

Even though curation and analysis work of the database is still ongoing, some
relevant facts about the biomedical research can be said. Over the period of 12
years, 32 general research areas have been defined, the three most funded
research areas are \emph{chronic and degenerative diseases}, \emph{malignant
  neoplasms}, and \emph{infectious and parasitic diseases}. The least funded
area is \emph{Ethics and medicine}. The area with the most researchers is
\emph{malignant neoplasms}. Other areas of relevance for M\'exico are
\emph{diseases related to poverty} and \emph{Health and vulnerable groups}. \\

@@

Besides the characteristics of the population there are some other boundary
conditions that play an important role on the network topology and dynamics,
that motivated the development of our model. Biomedical research in M\'exico
constitutes a vibrant community and collaboration is part of everyday
work. However, M\'exico does not have public biobanks for research purposes
(which are specially relevant for research in genomics, for example), there is
no regulation on the access to biological samples such as tissue, cells, DNA,
RNA, etc.\footnote{Regulation exists regarding researcher-subject relations
  based on legal and ethical grounds. Also, all projects need to be approved by
  the Ethics Committee and IRB.} Something similar happens with data. There have
been some attempts to create open data repositories for biomedical research, but
they have not been established yet. Regulation on these subjects is still
missing. Finally, technology (e.g., PCR, sequencing and expression profiling
technology) is in the domain of the institutions with the highest research
profiles and sometimes PIs see technology as a personal good.\\

From our ethnographic work to date, we have been able to see that
biological samples, data and technology can become instruments for negotiating
collaboration. For example, among people involved in research projects, there
are researchers that do not have direct access to samples, simply because their
institution does not offer clinical services. Many of them are non medical
doctors but chemists, biologists, physicists, and mathematicians. There is
another group of researchers that are placed on hospitals who are able to do
research and have access to biological samples from their own patients. It seems
that this group is the most privileged one, and the least pressured to establish
collaboration at whatever cost. Finally, there is one more group formed by those
who work as clinicians at small hospitals with no research infrastructure
whatsoever. This group may have an interest in research and the way for them to
become part of a project and be listed as authors in scientific papers is by
giving researchers who do not have access to biological samples access to
patients.\\

@@

Our model is based on the iterated version of the Prisoner's Dilemma (PD) and a
coordination game instantiated on
networks. Implementing games on networks is not new and it's an active area of
research aimed to understand the evolution of cooperation in networks populated
by selfish agents \cite{Szabo2007,Nowak1992,Nowak2006,Santos2005,Santos2006}. In many network models
on which some of game theory games are simulated, agents' decision to cooperate
or defect depend on a specific strategy, such as the well known
\textit{tit-for-tat} \cite{Axelrod2006,Nowak2011}. In some other cases, agents
can modify the weight of the interactions with their neighbors
\cite{Santos2006}. From a different perspective others have explored the effect
of different topologies on the emergence of cooperation
\cite{Santos2005,Hauert2004}. {\color{red}In our case, an agent's decision to cooperate or
defect is an outcome that depends on a balance between
utilities and the current strength of its collaboration relationships. Such
balance reflects the overall success or failure of its strategies}. We
study the behavior of the system under different topologies, including a
real-world network.\\

@@

The strength of the connection, represented by $w$, is a property of the link between two agents and
gets updated according to an $A_{ij}$ matrix of a coordination game. In the $w$ matrix, the highest value
goes to an edge when both agents cooperate, getting an $R$ for reward, if one of
them defects, the connection gets weaker getting $P$ for the collaborative
connection being punished. If both agents defect, the value $w$ doesn't change,
which means that agents didn't interact or that the interaction gets nullified
$N$. In this game, the best action for any agent is to coordinate with its
neighbor, either beacuse it wins or it doesn't loose. \\

@@

After each game, the agent adds-up utility ($u$), which is the sum of the
pay-offs following the PD matrix. A pair of neighbors will add-up to the strength of
their connection ($w_{ji}$) as they coordinate or anti-coordinate, being $w$
also cumulative. We measure global utility and connection strength for the whole
network. Global utility $U$ is the sum of all individual utilities and global
strength of connections or $W$ is the sum of every pair of agents' links
$w$. Metaphorically, the strength of connection can 
be thought as some sort of ``trust''.\\

{\color{red} It should be noted that the same actions or behaviors work for
  both $u$ and $w$. There are two reasons for this decision in the design of the
model. The most general one is that we believe that in the real world,
actions such as cooperating and defecting affect the strength of the connection
among people. The second one is that we think that selfishly maximizing resources and
strenghting relationships are \emph{opposing forces} acting on the same set of
behaviors. The actions of an agent imply a trade-off in which defecting may
increase its utility at the expense of its collaborative relationships. If
collaborators have nourished their relationships, they might be strong enough to
endure occasional defection. Cooperating may build up relationships but it can be
expensive for the player.}

@@

We ran two different experiments. In the first we simulated different values of
carrying capacity $\nu$. With this experiment we were able to see how the number of
cooperators, utility, strength of connections among agents and the ratio of
shifting state population would change in the range of the carrying
capacity. For this experiment the states of the agents were the same at
initialization, for all values of the carrying capacity. Since the model is
deterministic, it will return the same result if run under the same
conditions.\\

@@

Watts-Strogatz networks \cite{Watts1998} (small-world networks) are in a regime between a fully regular grid (lattice) and a random network (Erd\"{o}s-R\'enyi). In order to build them, a node is chosen from a lattice (a ring) and the edge that connects it to nearest neighbor in a clockwise sense. With probability $p$, this edge is reconnected to a node chosen uniformly at random over the entire ring, with duplicate edges forbidden; otherwise the edge is left in place. This process is repeated by moving clockwise around the ring, considering each node in turn until one lap is completed. Next, the edges connect nodes to their second-nearest neighbors clockwise. And as before, each of these edges is randomly rewired with probability $p$, and continue this process, circulating around the ring and proceeding outward to more distant neighbors after each lap, until, each edge in the original lattice has been considered once. The main characteristic of these networks is that the average shortest path length is small and grows as $log(N)$ ($\langle l \rangle \sim log(N)$). Also, the average clustering coefficient $\langle C \rangle$ remains large in terms of $p$. For $p < 0.1$, $\langle C \rangle \sim 1$.

@@

The biomedical research network on which we are running our model was generated 
with data from collaborative projects. Our data was obtained from CONACyT and includes information for twelve years of \textit{FOSISS} grants. Data included names of Principal Investigators, collaborators, research topics, etc. The network we are using here has researchers as nodes and edges represent the connection of two scientists when they colaborate in the same project. Edges are also weighted according to the number of projects shared by any pair of scientists.\\

@@

Other results to report are those of the dynamics of different
variables as the carrying capacity $\nu$ changes. The  most salient
result is that for $\nu$ between $0.19$ and $0.24$, there is an
apparent phase transition in all different topologies and for all the
different variables. Nevertheless, it is worth noticing that the shape
of the phase transition is different according to the topology of the
network at stake. When $\nu$ is between $0.0$ and $0.2$, that is, when there is
no space or a very short one for suspiciousness, all agents cooperate, when
\textit{carrying capacity} is above $0.25$, all agents defect. \textit{Utility},
\textit{strength of interactions}, and \textit{changing state population}
replicate that same behavior for the same limits. \\

@@

In the case of the Watts-Strogatz, small-world network, the whole population
remains cooperating for ranges between $0$ and $0.2$ but as it gets closer to
$0.2$, more time is needed for the population to become full of cooperators. In
$\nu \approx 0.2$ the cooperators will represent only half of the population and
such number of them will be constant up to $\nu \approx 0.25$ forming a short
plateau. From $\nu \approx 0.25$ to $\nu \approx 0.6$ cooperators will be
present at the beginning of the simulation but will go diminishing as time goes on.
In the case of the Barab\'asi-Albert network, crossing the threshold of $\nu
\approx 0.2$, there is a sharp decrease in the number of cooperators, but stays
constant over time. Such behavior is present for a very short range of $\nu$, and
before $\nu \approx 0.24$, cooperators disappear for the rest of values of
$\nu$. Finally, FOSISS network behaves similar to the other networks in that
there is a fall in the number of cooperators close  $\nu \approx 0.2$. Different
to the other networks, FOSISS network lacks the sharp reduction of
cooperators, instead this population declines smoothly and progressively;
specially, when it reaches a $\nu \approx 0.25$ cooperators go 
decreasing in a less dramatic manner all the way to $\nu \approx 0.5$. It is
also noteworthy that from $\nu = 0$ to $\nu \approx 0.5$ the number of
cooperators converge to a certain amount and stays constant for the rest of the
simulation.

\begin{figure} [h!]
\centering
\begin{tabular}{cc}
  \includegraphics[scale=0.28]{images/erdos_dc_nus_1.png} & \includegraphics[scale=0.28]{images/watts_dc_nus_1.png}
  \\
  \includegraphics[scale=0.28]{images/barabasi_dc_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_dc_nus_1.png}
\end{tabular}
\caption{Proportion of cooperators as a function of carrying capacity and time.}\label{CD}
\end{figure}

@@

Utility and strength of interactions dynamics under different $\nu$ emulate the
phase transition found before. In \textbf{Figures 3 and 4}, it can be observed
that there is a drop in utility and strength of interactions according to the
drop in the number of cooperators. Erd\"{o}s-Reny\'i and Barab\'asi-Albert
networks are quite similar in the way these variables fall in two steps, the
first one at $\nu \approx 0.2$ and the next one at $\nu \approx 0.23$. The fall
is sharper still in the Barab\'asi-Albert topology. Utility and strength of
interactions phase transition in Watts-Strogatz network is significantly more
staggered compared to the former networks. In the case of utility, there is a
region in the limits of $\nu \approx 0.25$ and $\nu \approx 0.3$, before utility
goes to $0$, in which it remains low but stable over time. In general, strength
of interactions follows the same pattern as utility but in the same $\nu \approx
0.25$ and $\nu \approx 0.3$, strength of interactions grows to a value that is
higher than the one given by default but soon starts to decrease as the
simulation runs. For FOSISS network, utility and strength of interactions fall
is quite steep but smooth, without sharp cuts. Between $\nu \approx 0.23$ and
$\nu \approx 0.28$, utility and strength of interactions start at their lowest,
but there is a slight increase in both of them.

@@

\includegraphics[scale=0.28]{images/erdos_state_nus.png} & \includegraphics[scale=0.28]{watts_state_nus.png} \\
\includegraphics[scale=0.28]{images/barabasi_state_nus.png} & \includegraphics[scale=0.28]{fosiss_state_nus.png}
\end{tabular}
\caption{Shifting population between cooperators and defectors as a function of
  carrying capacity $\nu$ and time.}\label{state} 
\end{figure}

>>>>

  What would happen if researchers in a community were given the same
  amount of resources and then set free to interact among each other?
  How would these resources be distributed in the population after
  many interactions? The final result may be determined by the
  researchers' social and political power, an by their
  prestige. Nevertheless, all these attributes depend, to a certain
  degree, on the research collaboration social network in which
  scientists are embedded. Here we develop a model based on the
  Prisoner's Dilemma implemented on a collaboration network. With this
  model we explore the distribution of resources and the dynamics of
  the strengthening or weakening of collaboration interactions among
  partners, as scientists cooperate or defect in order to gain access
  to resources shared by others. The network is based on data about projects
  that were awarded with a grant. In this sense, we assume that
  resources are funds, equipment and time. We define Fitness as how
  much access a researcher has to the resources of his collegues.  We
  tested our simulation on the real biomedical research network and
  compared the results with an Erd\"{o}s-Reny\'i, a Watts-Strogatz
  small-world and Barab\'asi-Albert topologies. Different topologies
  display different fitness and connections strength
  distributions. Moreover, the distribution of fitness and connections
  strength in the researchers network is similar to that of
  Barab\'asi-Albert and Watts-Strogatz topologies, respectively. We
  believe that fitness distribution in the researchers network
  suggests that there are socio-cultural mechanisms governing the
  network that produce an asymmetric distribution of resources. The
  high distribution of strong connections might reflect some sort of
  subordination among researchers by which they are morally obliged to
  cooperate by the same socio-cultural mechanisms. The range around
  the threshold that regulates the decision to cooperate or defect
  according to the agent's historical balance between utility and
  strength of collaborative relationships and carrying capacity of the
  system is small, suggesting that there is a region in which a phase
  transition takes place from a population of cooperators to a
  population of defectors. Simulations like this may help to develop
  science policies to promote fair distribution of resources.

@@

%% \textbf{Notes:}

%% \begin{itemize}

%% %% \item In this paper there are two games played simultaneously: One for
%% %% maximazing individual utility; the second for maximazing the connection
%% %% strength. We are interested in how they affect eachother in the context of a
%% %% network of collaboration.

%% \item We change utility for fitness. Utility seems a better concept in order to
%%   reflect players' preferences.

%% %% \item In the context of our paper, utility represents
%% %%   the preference for having access to the resources of others. Maximazing
%% %%   utility means that the player is increasing its access to the resources of
%% %%   other players.  Utility function is just the sum of the payoffs of playing
%% %%   with its neighbors. As for the payoff matrix, the best strategy for a player
%% %%   is to defect. Any other option will not be an optimal payoff. The opossing
%% %%   force is in the other game. The game for maximazing the strength of the
%% %%   connection among players. In this case defecting or cooperating? is the best
%% %%   strategy as it pays the most when coordinating regardeless they both cooperate
%% %%   or defect, this is because when both cooperate the interaction gets a positive
%% %%   payoff, when both defect, the interaction doesn't get affected; but if they
%% %%   anti-coordinate, then the interaction looses. In this sense, this might be a
%% %%   game were there are multiple (i.e., 2 points or strategies) equilibria. It
%% %%   seems that there is no single dominant strategy.}\\

%% \item Choosing a strategy depends on the balance $\eta'$ equation --look at page
%%   9. This
%%   equation states that the probability of an agent to cooperate increases as: a)
%%   its utility increases --as it moves away from 0, or; b) The strength of the
%%   connection with its peers increases --as it moves away from 0, or; c) both
%%   increases.

%%   $\eta_i = \frac{\langle f_i \rangle + \langle w_{ij} \rangle _j}{2}$\\

%% \item{En la pagina 9 decimos que ``In order for the agent to choose to cooperate
%%   or defect, a degree of willingness to cooperate is assigned to each agent.'' No estoy
%%   seguro que ``degree of willingness to cooperate'' sea la mejor manera de
%%   decirlo, mi duda está en la palabra ``degree'', por lo que cambi\'e la
%%   oraci\'on a: The probability for an agent to cooperate or defect depends on a
%% number that referes to a historical balance between average utility and the average strength of
%% the connections with its neigbors. }

%% \end{itemize}

@@

 {\color{red}In this article we explore the network effect on the distribution
   of players having access to certain amount of resources shared by other players in the
   network and the distribution of the strength of interactions among
   them. Particularly, we implemented two games played simultaneously: 
   one for maximazing individual utility based on the iterated Prisoner's
   Dilemma; the other, a coordination game for maximazing the connection strength between
   players. We are interested in how they affect each other in 
   the context of a network of scientific collaboration under the idea that while
   researchers are interested in maximazing their utilities, they also know that
   it is important to invest in building collaborative relationships. These two
   behaviors are explored in a biomedical research community of M\'exico.}\\

{\color{red}In the context of our paper, utility represents
  access to resources shared by others.
  The value of the Utility function for a player is the sum of the payoffs of playing
  with its neighbors. The opossing force comes from the other concurrent game: players
  trying to maximize the strength of their interactions with other players. In the coordinating game
  the best strategy is to adopt the same strategy as the other player, as it pays the most
  regardless of cooperation or defection in the utility game. When
  both cooperate the interaction gets a positive payoff, when both defect, the
  interaction doesn't get affected; but if they anti-coordinate, then the
  interaction looses. Finally, cooperation is a central feature of scientific work.
  For our biomedical network, cooperation can be thought of
  as sharing resources such as time, students, equipment, even money. Examples of defection
  to a cooperator are ghost authorship or
  prestige authorship.

  % TODO: este ejemplo de defection no es nada claro:
  %
  % Another example may be a working strategy in which some
  % members of a team do the work at certain moment, and then another part 
  % does its part in some other moment in the future --some sort of
  % reciprocity.

@@

The manuscript is structured in five sections. First we describe
\textit{FOSISS}, the main program for grants destined to biomedical applied
research in M\'exico. This is the source of the
database from which we created the researchers collaboration network. Next we
describe our model
and the different network topologies on which we explored it.
We then present our results and discuss them. In the last section
we draw some final remarks and conclusions.

@@

\emph{FOSISS} or Sectoral Fund for Health and Social Security Research
(\emph{Fondo Sectorial en Investigaci\'on en Salud y Seguridad Social}) is one of
such funds. FOSISS is constituted by CONACyT, SSA, IMSS and ISSSTE,\footnote{SSA
  is the acronym for Secretariat of Health \emph{Secretar\'ia de Salud}; IMSS is
  the acronym for Social Security Mexican Institute (\emph{Instituto Mexicano
    del Seguro Social}); ISSSTE stands for Institute for Social Security and
  Services for State Workers (\emph{Instituto de Seguridad y Servicios Sociales
    de los Trabajadores del Estado})} all of them being the major public health
providers and research institutions in the country. Every year CONACyT opens a
call for funds limited to a set of health research areas previously defined by a
group of experts. Such areas range from public health issues to chronic and
degenerative diseases.\\

Most applicants are public universities and research institutions, but
eligibility is open to public and private health research sectors.
From 2002 through 2013, there were 91 institutions funded that
comprised 4988 researchers.\\

From these data some important considerations should be made
clear. Scientists in the database take on the roles of principal
investigators (PIs), associate researchers, postdoctoral associates,
postgraduate and undergraduate students.
Unfortunately, information on these roles is not specified in the database.
We acknowledge the importance of this deficiency because
researchers in our network act under different circumstances and we know
that this diversity has a real impact on the structure and eventually on the
dynamics of the network, as well as on the results of our model.\\

Our database includes the
name of the project, the year it was approved for funding and the research area
to which it was assigned. It specifies the names of PIs or the people 
responsible for the project and the names of collaborators.
Researchers can be PIs in one project and collaborators on a different project. The institutional
affiliation of all participants is included. Through this affiliation
we determine the principal institution
behind every project.\\

Even though curation and analysis work of this database is still going on, some
relevant facts about the biomedical research can be said. Over the period of 12
years, 32 general research areas have been defined, the three most funded
research areas are \emph{chronic and degenerative diseases}, \emph{malignant
  neoplasms}, and \emph{infectious and parasitic diseases}. The least funded
area is \emph{Ethics and medicine}. The area with the most researchers is
\emph{malignant neoplasms}. Other areas of relevance for M\'exico are
\emph{diseases related to poverty} and \emph{Health and vulnerable groups}. \\

@@

Besides the characteristics of the population there are some other boundary
conditions that play an important role on the network topology and dynamics,
that motivated the development of our model. Biomedical research in M\'exico
constitutes a vibrant community and collaboration is part of everyday
work. However, M\'exico does not have public biobanks for research purposes
(which are specially relevant for research in genomics, for example), there is
no regulation on the access to biological samples such as tissue, cells, DNA,
RNA, etc.\footnote{Regulation exists regarding researcher-subject relations
  based on legal and ethical grounds. Also, all projects need to be approved by
  the Ethics Committee and IRB.} Something similar happens with data. There have
been some attempts to create open data repositories for biomedical research, but
they have not been established yet. Regulation on these subjects is still
missing. Finally, technologically advanced equipment such as high throughput sequencers
are kept by institutions with the highest research
profiles and sometimes PIs manage them in a self serving way.\\

From our ethnographic work to date, we have been able to see that
biological samples, data and technology can become instruments for negotiating
collaboration. For example, among people involved in research projects, there
are researchers that do not have direct access to samples, simply because their
parent institution does not offer clinical services. Many of them are non medical
doctors but chemists, biologists, physicists, and mathematicians. There is
another group of researchers that are placed on hospitals which is able to do
research and have access to biological samples from their own patients. It seems
that this group is the most privileged one, and the one with the least pressure to establish
collaboration at whatever cost. Finally, there is one more group formed by those
who work as clinicians at small hospitals with no research infrastructure
whatsoever. This group may have an interest in research and the way for them to
become part of a project and be listed as authors in scientific papers is by
giving researchers who do not have access to biological samples access to
patients.\\

@@

Our model is based on the iterated version of the Prisoner's Dilemma (PD) and a
coordination game instantiated on
networks. Implementing games on networks is not new and it's an active area of
research aimed to understand the evolution of cooperation in networks populated
by selfish agents \cite{Szabo2007,Nowak1992,Nowak2006,Santos2005,Santos2006}. In many network models
on which some of game theory games are simulated, agents' decision to cooperate
or defect depend on a specific strategy, such as the well known
\textit{tit-for-tat} \cite{Axelrod2006,Nowak2011}. In some other cases, agents
can modify the weight of the interactions with their neighbors
\cite{Santos2006}. From a different perspective others have explored the effect
of different topologies on the emergence of cooperation
\cite{Santos2005,Hauert2004}. {\color{red}In our model, an agent's decision to cooperate or
defect depends on a balance between
utilities and the current strength of its collaboration relationships. Such
balance reflects the overall success or failure of its strategies}. We
study the behavior of the system under different topologies, including a
real-world network.\\

@@

The strength of the interaction, represented by $w$, is a property of the link between two agents and
gets updated according to an $A_{ij}$ matrix of a coordination game. In the $w$ matrix, the highest value
goes to an edge when both agents cooperate, getting an $R$ for reward, if one of
them defects, the connection gets weaker getting $P$ for the collaborative
connection being punished. If both agents defect, the value $w$ doesn't change,
which means that agents didn't interact or that the interaction gets nullified
$N$. In this game, the best action for any agent is to coordinate with its
neighbor, either beacuse it wins or because it doesn't loose. \\

@@

After each game, the agent adds-up utility ($u$), which is the sum of the
pay-offs following the PD matrix. A pair of neighbors will add-up to the strength of
their interaction ($w_{ji}$) as they coordinate or anti-coordinate, being $w$
also cumulative. We measure global utility and connection strength for the whole
network. Global utility $U$ is the sum of all individual utilities and global
strength of connections or $W$ is the sum of every pair of agents' links
$w$. The strength of interaction can 
be thought as some sort of ``trust''.\\

{\color{red} It should be noted that the same actions or behaviors work for
  both $u$ and $w$. There are two reasons for this decision in the design of the
model. The most general one is that we believe that in the real world,
actions such as cooperating and defecting affect the strength of the interaction
among people. The second one is that we think that selfishly maximizing access to resources and
strengthening relationships are \emph{opposing forces} acting on the same set of
behaviors. The actions of an agent imply a trade-off in which defecting may
increase its utility at the expense of its collaborative relationships. If
collaborators have nourished their relationships, they might be strong enough to
endure occasional defection. Cooperating may build up relationships but it can be
expensive for the player.}

@@

We ran two different experiments. In the first we simulated different
values of carrying capacity $\nu$. With this experiment we were able
to see how the number of cooperators, utility, strength of connections
among agents and the ratio of shifting state population would change
in the range of the carrying capacity. The states of the agents were
the same at initialization, for all values of the carrying
capacity. Since the model is deterministic, it will return the same
result if run under the same conditions.\\

@@

Watts-Strogatz networks \cite{Watts1998} (small-world networks) are in a regime between a fully regular grid (lattice) and a random network (Erd\"{o}s-R\'enyi). In order to build them, a node is chosen from a lattice (a ring) and the edge that connects it to nearest neighbor in a clockwise sense. With probability $p$, this edge is reconnected to a node chosen uniformly at random over the entire ring, with duplicate edges forbidden; otherwise the edge is left in place. This process is repeated by moving clockwise around the ring, considering each node in turn until one lap is completed. Next, the edges connect nodes to their second-nearest neighbors clockwise. As before, each of these edges is randomly rewired with probability $p$. This process continues, circulating around the ring and proceeding outward to more distant neighbors after each lap, until each edge in the original lattice has been considered once. The main characteristic of these networks is that the average shortest path length is small and grows as $log(N)$ ($\langle l \rangle \sim log(N)$). Also, the average clustering coefficient $\langle C \rangle$ remains large in terms of $p$. For $p < 0.1$, $\langle C \rangle \sim 1$.

@@

The biomedical research network on which we are running our model was generated 
with data from collaborative projects. Our data was obtained from CONACyT and includes information for twelve years of \textit{FOSISS} grants. Data included names of Principal Investigators, collaborators, research topics, etc. The network we are using here has researchers as nodes and edges represent the connection of two scientists when they colaborate on the same project. Edges are also weighted according to the number of projects shared by any pair of scientists.\\

@@

Other results to report are those of the dynamics of different
variables as the carrying capacity $\nu$ changes. The  most salient
result is that for $\nu$ between $0.19$ and $0.24$, there is an
apparent phase transition in all different topologies and for all the
different variables. It is worth noting that the shape
of the phase transition is different according to the topology of the
network under study. When $\nu$ is between $0.0$ and $0.2$, that is, when there is
no space or a very short one for suspiciousness, all agents cooperate, when
\textit{carrying capacity} is above $0.25$, all agents defect. \textit{Utility},
\textit{strength of interactions}, and \textit{changing state population}
replicate that same behavior for the same limits. \\

@@

In the case of the Watts-Strogatz, small-world network, the whole
population remains cooperating for ranges between $0$ and $0.2$ but as
it gets closer to $0.2$, more time is needed for the population to
become full of cooperators. In $\nu \approx 0.2$ the cooperators will
represent only half of the population and such number of them will be
constant up to $\nu \approx 0.25$ forming a short plateau. From $\nu
\approx 0.25$ to $\nu \approx 0.6$ cooperators will be present at the
beginning of the simulation but will diminish as time goes on.  In the
case of the Barab\'asi-Albert network, crossing the threshold of $\nu
\approx 0.2$, there is a sharp decrease in the number of cooperators,
but stays constant over time. Such behavior is present for a very
short range of $\nu$, and before $\nu \approx 0.24$, cooperators
disappear for the rest of values of $\nu$. Finally, FOSISS network
behaves similarly to the other networks in that there is a fall in the
number of cooperators close to $\nu \approx 0.2$. In contrast with the
other networks, the FOSISS network lacks the sharp reduction of
cooperators, instead this population declines smoothly and
progressively; specially, when it reaches a $\nu \approx 0.25$
cooperators decrease in a less dramatic manner all the way to
$\nu \approx 0.5$. It is also noteworthy that from $\nu = 0$ to $\nu
\approx 0.5$ the number of cooperators converge to a certain degree
and stays constant for the rest of the simulation.

\begin{figure} [h!]
\centering
\begin{tabular}{cc}
  \includegraphics[scale=0.28]{images/erdos_dc_nus_1.png} & \includegraphics[scale=0.28]{images/watts_dc_nus_1.png}
  \\
  \includegraphics[scale=0.28]{images/barabasi_dc_nus_1.png} & \includegraphics[scale=0.28]{images/fosiss_dc_nus_1.png}
\end{tabular}
\caption{Ratio of cooperators as a function of carrying capacity and time. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.
}\label{CD}
\end{figure}

@@

Utility and strength of interactions dynamics under different $\nu$ are similar to the
phase transition found before. \textbf{Figures 3 and 4} show
that there is a drop in utility and strength of interactions according to the
drop in the number of cooperators. Erd\"{o}s-Reny\'i and Barab\'asi-Albert
networks are quite similar in the way these variables fall in two steps, the
first one at $\nu \approx 0.2$ and the next one at $\nu \approx 0.23$. The fall
is sharper still in the Barab\'asi-Albert topology. Utility and strength of
interactions phase transition in Watts-Strogatz network is significantly more
staggered compared to the former networks. In the case of utility, there is a
region in the limits of $\nu \approx 0.25$ and $\nu \approx 0.3$, before utility
goes to $0$, in which it remains low but stable over time. In general, strength
of interactions follows the same pattern as utility but in the same $\nu \approx
0.25$ and $\nu \approx 0.3$, strength of interactions grows to a value that is
higher than the one given by default but soon starts to decrease as the
simulation runs. For the FOSISS network, utility and interaction strength fall
quite steeply but smoothly, without sharp cuts. Between $\nu \approx 0.23$ and
$\nu \approx 0.28$, utility and strength of interactions start at their lowest,
but there is a slight increase in both of them.

@@

\includegraphics[scale=0.28]{images/erdos_state_nus.png} & \includegraphics[scale=0.28]{images/watts_state_nus.png} \\
\includegraphics[scale=0.28]{images/barabasi_state_nus.png} & \includegraphics[scale=0.28]{images/fosiss_state_nus.png}
\end{tabular}
\caption{Shifting population between cooperators and defectors as a function of
  carrying capacity $\nu$ and time. Topologies are: A. Random, Erd\"{o}s-R\'enyi. B. Watts-Strogatz. C. Barab\'asi-Albert. D. FOSISS.}\label{state} 
\end{figure}
